\chapter{Example Auxiliary Files}\label{ChapAppendAuxFiles}
\section{Ash3d Control File examples}\label{ChapAppendSecInputFile}
\small
\begin{verbatim}
*******************************************************************************
1992 08 19   1.0   3.5     13.7   0.014
*******************************************************************************
\end{verbatim}
\normalsize

\section{Airport/Point-of-Interest examples}\label{ChapAppendSecPOI}
Ash3d can read an external file that lists Airports and/or points of interest (POI).
An example of the format for this file can be seen in the global list of
6117 airports included in the \texttt{volcano-ash3d} repository in
\texttt{GlobalAirports\_ewert.txt}. This file is copied to the install
location \texttt{INSTALLDIR/share} where it is read at run-time if Ash3d
is compiled with \texttt{USEEXTDATA = T} in the makefile. The file has the
following format. The first line is a header line describing the columns.
All following lines must contain the coordinates of the point-of-interest in
the first 50 characters of each line.
The latitude and longitude values are populated first with the first two columns
of data, then the x,y projected values if columns 3 and 4 are present.
Characters 51-53 contain the 3-char ICAO airport code. Characters 55-89 contain
the station name, which is usually the airport name with the state and/or
country.

Here is the top few lines of \texttt{GlobalAirports\_ewert.txt} as an example.
\footnotesize
\begin{verbatim}
[volcano-ash3d]$ head share/GlobalAirports_ewert.txt 
#   Latitude   Longitude           x           y  Code Location
    41.60833   -88.09417                          II2  Lewis University Airport, IL
    34.68861   -85.29056                          GA1  Barwick Lafayette Airport, GA
    44.25083   -90.85528                          WI1  Black River Falls Area Airport, WI
    50.40000   136.95000                          KXK  Kimsomolsk Na Amure, Russia (European)
    58.36528  -152.69667                          A33  Hidden Lake, AK
    57.51139  -152.95278                          A32  Hidden Basin, AK
    58.36194  -152.31222                          A34  Laura Lake, AK
    57.64222  -153.02194                          A30  Terror Lake, AK
    57.32306  -153.14944                          A29  Kiluda Bay, AK
\end{verbatim}
\normalsize

\small
\begin{verbatim}
#   Latitude   Longitude           x           y  Code Location
    46.01217  -121.90125                               ASW6_24_1
    46.05025  -121.91429                               ASW6_24_2
    46.06452  -121.98259                               ASW6_24_3
\end{verbatim}
\normalsize

MAXAIRPORTS = 10000
6117

\section{Volcano file}
VotW\_ESP\_v12\_csv.txt


